---
title: "Homework 1"
output: html_document
---

```{r include = FALSE}
# This code block sets up the r session when the page is rendered to html
# include = FALSE means that it will not be included in the html document

# Write every code block to the html document 
knitr::opts_chunk$set(echo = TRUE)

# Write the results of every code block to the html document 
knitr::opts_chunk$set(eval = TRUE)

# # Define the directory where images generated by knit will be saved
# knitr::opts_chunk$set(fig.path = "images/03-A/")

# Set the web address where R will look for files from this repository
# Do not change this address
repo_url <- "https://raw.githubusercontent.com/LivingLandscapes/Course_EcologicalModeling/master/"

# Suppress warnings and messages
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

```

## Overview

Please provide a reproducible .R script that answers all questions below and produces all of your analysis and plots. Some reminders:

- **Reproducible:** This means when I run your .R file, it will run all analyses and create all plots without errors, without me having to reset my working directory, and without forcing me to install anything on my machine (i.e., use the require(librarian); shelf(your packages, lib = tempdir()) approach as in the code block below). This also means there should be *NO ERRORS* when I run it!
- **Format:** Answer each question with (commented) text and code.

### Background

The [Valles Caldera National Preserve](https://www.nps.gov/vall/index.htm) is part of the National Parks. It's the caldera of an old volcano in the Jemez Mountains of New Mexico, USA. It has some pretty interesting socio-ecological history. Part of that history is CRAZY overgrazing by sheep and cattle during Spanish and then American ownerships. This overgrazing resulted in massive plant biodiversity losses in the native grasslands and riparian areas in the Caldera. In the early 2000's, some scientists established a grazing exclosure experiment along several of the riparian zones in the Caldera. Their goal was to see if excluding cattle or all ungulates (cattle, deer, and elk) would increase plant biodiversity in the riparian areas.

Some details on the data:

| **Column Name**| **Description**                                             |
|----------------|-------------------------------------------------------------|
| year             | years 2002 - 2014                                          |
| site             | unique code for each site/treatment |
| treatment        | treatment type. C = control, LE = livestock exclosure, UE = all ungulates exclosure                  |
| All other columns|  each column is a plant species, and values are number of observations per 300 points within each site.                                                   
--- 

## Your assignment

1) The `site` metadata column has too much information: we need to split the information up into separate columns of it out while  retaining the unique identifier information. Revise the `site` column to be the first three characters in the current `site` column (e.g., the new site value for RS1C1 should be "RS1"), and create a new column named `plotCode` with the final 1 - 2 characters in the current `site` column (e.g., the plotCode value for RS1C1 should be "C1"). To be clear: when you're finished with this problem, the metadata columns should be: `year`, `site`, `plotCode`, and `treatment`.

2) Create a `Richness_total` column by calculating total species richness for each row, and create an `Diversity_InvSimp` column by calculating Inverse Simpson's Diversity Index for each row. Hint: for the Inverse Simpson's, check out the `vegan::diversity` function. Also, total species richness is the total number of species per row.

<!-- 3) Use the wetland obligate species list (see code block below) to create a column of *wetland obligate*  species richness. That is, the number of wetland obligate species per row. -->

3) Using boxplots and histograms, check for outliers in the total species richness and species diversity columns.

4) Check for balance in the data across sites, years, and treatments. Hint: use the `table` function. Using commented text, say if data is balanced and if not, how so?

5) Create plot(s) to check for relationships between response variables (richness, diversity) and predictor variables (year, site, treatment). Using commented text, describe potential relationships.

6) Why or why not should we worry about collinearity in our predictor variables?

7) Create plot(s) to check for a potential interaction between `year` and `treatment`. Using commented text, tell me why we should or shouldn't consider this interaction?

``` {r, include = TRUE, eval = FALSE}

# List of packages necessary to run this script:
require(librarian, quietly = TRUE)
shelf(tidyverse, sf, vegan, quiet = TRUE)

# Repo URL 
repo_url <- "https://raw.githubusercontent.com/LivingLandscapes/Course_EcologicalModeling/master/"

# Load riparian vegetation monitoring dataset:
dat <- 
  read_csv(paste0(repo_url, "data/rip.dat.02-14.csv"))

```


<!-- # # Wetland obligate species of interest -->
<!-- # wetObl <-  -->
<!-- #   c("ALAE", "CAAQ",	"CAPEL",	"CASIM",	"CAUT",	"CIMA",	"ELAC",	"ELPAL",	 -->
<!-- #     "GLGR",	"GLST",	"LEMI", "MIGL",	"POAN",	"PONA",	"RAMA",	"SPAN",	"TOPA") -->